---
title: "Machine Learning Project"
output: pdf_document
---

**Name**: Sai Jaswanth Kumar Kunku  
**G Number** : G01331645

# Library Import
```{r warning = FALSE, message = FALSE}
# Suppress dplyr summarise grouping warning messages
options(dplyr.summarise.inform = FALSE)

library(tidyverse)
library(tidymodels)
library(vip)
library(discrim)
library(ggcorrplot)
library(rpart.plot)
library(psych)
library(gridExtra)
```
# Raw Data

```{r}
credit_card_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/credit_card_df.rds'))
#write.csv(credit_card_df, "credit_card_data.csv", row.names=FALSE)
summary(credit_card_df)
```


# Question 1: What is the distribution of open and closed accounts for different age groups?

**Answer**:The frequency of customers who are closing their accounts are high in the age groups between 35-44 and 45-54 respectively.This is the age group that has more number of active accounts as well.

```{r}
credit_card_df <- credit_card_df %>% 
  mutate(age_group = case_when(
    age >= 25 & age <34 ~ '25-34',
    age >= 35  & age <44 ~ '35-44',
    age >= 45 & age <54 ~ '45-54',
    age >= 55 & age <64 ~ '55-64',
    TRUE ~ '64 and older'))

df1<-credit_card_df %>% group_by(age_group,customer_status) %>% 
summarize(total_count=n()) %>%
mutate(percentage =  round(100*(total_count/sum(total_count)),2))

ggplot(df1,aes(x=age_group,y=total_count,fill=customer_status))+
  geom_bar(stat="identity",color="black",position = "dodge") +
  labs(title = "Account Status distribution by Age Group",
             x = "Age Group",
             y = "Count")
  
  
```


# Question 2:How the income levels of different customers vary by their employment status and marital status?

**Answer**:The median income of full-time,part-time and self-employees customers lies between 40000 to 50000 and there is no noticeable differences in active vs closed accounts for married and unmarried status. 

```{r}


fig1<-credit_card_df %>% ggplot(aes(x=employment_status,
                         y=income,
                         fill=customer_status)) +
geom_boxplot()  +  labs(title = "Summary Statistics of Income by Employement Type", y = "Income",
         x = "Employement Status") 

fig2<-credit_card_df %>% ggplot(aes(x=marital_status,
                         y=income,
                         fill=customer_status)) +
geom_boxplot()  +  labs(title = "Summary Statistics of Income by Marital Status", y = "Income",
         x = "Marital Status") 

grid.arrange(fig1, fig2)
```


# Question 3: What percentage of active and churned customers with respect to their employment status

**Answer**:The results from pie chart shows that 90% of closed accounts are from part-time and full-time employees.It should be noticed that the proportion of closed accounts is 28% higher than that of corresponding active accounts.
```{r}
emp_data<-credit_card_df %>% group_by(customer_status,employment_status) %>% summarize(count=n()) %>%  mutate(percentage =  round((count*100/sum(count)),1))

emp_data

ggplot(emp_data, aes(x = "", y = percentage, fill = employment_status)) +
  geom_col(color = "black") +guides(fill=guide_legend(title="Employement Type"))+
  geom_text(aes(label = percentage),
            position = position_stack(vjust = 0.5)) +
  coord_polar(theta = "y")+theme_void() +  scale_fill_brewer() +facet_wrap(~customer_status)+
  ggtitle("Proportion of Active and Closed Accounts by Employee Type")+
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom",panel.spacing = unit(1, "lines"))
```


# Question 4: How does the Transaction ratio,number of transactions and spend ratio vary among different customers?

**Answer**:The average transaction ratio of active accounts is 0.74 whereas for the closed accounts is 0.56.The spending ratio of both active and closed accounts is almost identical with values at 0.77 and 0.7 respectively.
```{r}
ratio_data<-credit_card_df %>% 
  group_by(customer_status) %>%
  summarise(avg_transaction_ratio = round(mean(transaction_ratio_q4_q1),2),
            total_transactions = round(median(transactions_last_year),2),
            avg_spend_ratio = round(mean(spend_ratio_q4_q1),2)) 

ratio_data

outlier1 <- boxplot(credit_card_df$transaction_ratio_q4_q1, plot=FALSE)$out
x<- as.data.frame(credit_card_df[-which(credit_card_df$transaction_ratio_q4_q1 %in% outlier1),])

x %>% 
  pivot_longer(cols = c(transaction_ratio_q4_q1, utilization_ratio,spend_ratio_q4_q1), 
               names_to = 'value_type', 
               values_to = 'Ratio') %>%  ggplot(aes(x=value_type,
               y=Ratio,
               fill=customer_status)) +
  geom_boxplot()+ 
  labs(title = "Banking Pattern of Active vs Churned Customers",
       x = "Account Status")


```




# Question 5: Does no:of dependents has any impact on the status of the customer? 

**Answer**:The increase in number of dependents doesn't show any impact on the customer status as the percentage of closed accounts is always 10-15% lesser than active accounts within each dependent group.

```{r}
credit_card_df %>% group_by(dependents,customer_status) %>% summarize(count=n()) %>%  mutate(percentage =  round(100*(count/sum(count)),2)) %>% arrange(dependents)  
```
# Question 6:Which credit card holders are more likely to close their accounts and what is their average credit utilization ratio?

**Answer**:Blue card holders are more likely to close their accounts as they constitute 71.5% of overall closed accounts.Out of total blue card holders 58.6% has closed their accounts whereas followed by gold and silver card users with 32.9% and 25.3% respectively.


```{r}
credit_card_df %>% 
  group_by(card_type,customer_status) %>% summarize(count=n(),avg_utilization=mean(utilization_ratio)) %>% 
  mutate(percentage=round((count/sum(count))*100,2)) %>% arrange(customer_status
) %>% ungroup() %>% group_by(customer_status) %>%
  mutate(prop=round((count/sum(count))*100,2)) 
```


# Question 7::How are various characterstics related to each other?

**Answer**:There is positive correlation of 0.59 between the customer income and the credit limit.In addition to that the number of transactions made last year is positively related to the amount spent last year with a correlation of 0.80.On the flipside,credit limit and utilization ratio has negative correaltion of 0.39.

```{r,fig.width=7.5, fig.height=6}
numeric_df<-Filter(is.numeric, credit_card_df)
pairs.panels(numeric_df)
```


# Machine Learning

# Factoring & Data Splitting
```{r}
credit_card_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/credit_card_df.rds'))
credit_card_df$total_accounts<-factor(credit_card_df$total_accounts)
credit_card_df$contacted_last_year  <-factor(credit_card_df$total_accounts)
credit_card_df$months_inactive_last_year  <-factor(credit_card_df$months_inactive_last_year)
credit_card_df$years_active<-as.factor(round(credit_card_df$months_since_first_account/12))
credit_card_df$dependents<-factor(credit_card_df$dependents)
credit_card_df <- credit_card_df %>% 
  mutate(total_transactions = case_when(
    transactions_last_year <=55 ~ '<55',
    TRUE ~ '>55'))
credit_card_df<-credit_card_df %>% dplyr::select(-c("months_since_first_account","transactions_last_year","age"))

set.seed(1)
#Test/Train Split
credit_split <- initial_split(credit_card_df, prop = 0.75,strata = customer_status)
credit_training <- credit_split %>% training()
credit_test <- credit_split %>% testing()

set.seed(10)
#Cross Validation
credit_folds <- vfold_cv(credit_training, v = 5)
#Metrics
my_metrics <- metric_set(accuracy, sens, spec, f_meas, roc_auc)
```

# Feature Engineering
```{r}
credit_recipe <- recipe(customer_status ~ ., data = credit_training) %>% 
                   step_YeoJohnson(all_numeric(), -all_outcomes()) %>% 
                   step_normalize(all_numeric(), -all_outcomes()) %>% 
                   step_dummy(all_nominal(), -all_outcomes())

credit_recipe %>% 
  prep(training = credit_training) %>% 
  bake(new_data = NULL)
```



# Model 1 :Logistic Regression

```{r}
#Model Spec
logistic_model <- logistic_reg() %>% 
                  set_engine('glm') %>% 
                  set_mode('classification')

credit_wf <- workflow() %>% 
            add_model(logistic_model) %>% 
            add_recipe(credit_recipe)

credit_logistic_fit <- credit_wf %>% 
                      fit(data = credit_training)

credit_trained_model <- credit_logistic_fit %>% 
                       extract_fit_parsnip()

vip(credit_trained_model)

last_fit_model <- credit_wf %>% 
                  last_fit(split = credit_split,
                           metrics = my_metrics)

metrics<-last_fit_model %>% collect_metrics()

metrics

last_fit_results <- last_fit_model %>%  collect_predictions()

last_fit_results

last_fit_results %>% 
  roc_curve(truth = customer_status, estimate = .pred_closed_account) %>% 
  autoplot()

conf_mat(last_fit_results, truth = customer_status, estimate = .pred_class)
```


# Model 2 :KNN

```{r}


knn_model <- nearest_neighbor(neighbors = tune()) %>% 
             set_engine('kknn') %>% 
             set_mode('classification')

knn_wf <- workflow() %>% 
          add_model(knn_model) %>% 
          add_recipe(credit_recipe)

k_grid <- tibble(neighbors = c(10, 20, 30, 50, 70, 100))

## Tune  workflow
set.seed(10)

knn_tuning <- knn_wf %>% 
              tune_grid(resamples = credit_folds,
                         grid = k_grid)

knn_tuning %>% collect_metrics()

best_k <- knn_tuning %>% select_best(metric = 'roc_auc')

final_knn_wf <- knn_wf %>% finalize_workflow(best_k)

last_fit_knn <- final_knn_wf %>% last_fit(split = credit_split,metrics=my_metrics)

last_fit_knn %>% collect_metrics()

knn_predictions <- last_fit_knn %>% collect_predictions()

knn_predictions

knn_predictions %>% 
  roc_curve(truth = customer_status, estimate = .pred_closed_account) %>% 
  autoplot()

conf_mat(knn_predictions, truth = customer_status, estimate = .pred_class)
```


# Model 3 :Random Forest
```{r}
rf_model <- rand_forest(mtry = tune(),
                        trees = tune(),
                        min_n = tune()) %>% 
            set_engine('ranger', importance = "impurity") %>% 
            set_mode('classification')

rf_workflow <- workflow() %>% 
               add_model(rf_model) %>% 
               add_recipe(credit_recipe)

set.seed(10)

rf_grid <- grid_random(mtry() %>% range_set(c(2, round(sqrt(ncol(credit_training))))),
                       trees(),
                       min_n(),
                       size = 9)

set.seed(10)

rf_tuning <- rf_workflow %>% 
             tune_grid(resamples = credit_folds,
                       grid = rf_grid)

best_rf <- rf_tuning %>% 
           select_best(metric = 'roc_auc')

final_rf_workflow <- rf_workflow %>% 
                     finalize_workflow(best_rf)

rf_last_fit <- final_rf_workflow %>% 
               last_fit(split = credit_split,metrics=my_metrics)

credit_rf_fit <- final_rf_workflow %>% 
                      fit(data = credit_training)

rf_trained_model <- credit_rf_fit %>% 
                       extract_fit_parsnip()

vip(rf_trained_model)

metrics<-rf_last_fit %>% collect_metrics()

metrics

rf_last_fit %>% collect_predictions() %>% 
                roc_curve(truth = customer_status, estimate = .pred_closed_account) %>% 
  autoplot()

rf_last_fit %>% collect_predictions() %>% conf_mat(truth = customer_status, estimate = .pred_class)
```



# Summary of Results

# Introduction

Banking sector is one of the important sectors in financial paradigm as it plays an important role in our everyday life and in achieving economic growth.The number of customers closing their credit account is a concerning issue to banks as it affects their business and impacts overall profitability.Hence it is important for them to find effective ways to retain existing customers and prevent them from closing their accounts.The goal of this analysis to identify the key factors that are driving the customers to close their accounts,predict the likelihood of them and provide recommendations to prevent it from happening in the future.    

# Key Findings

Some of the key findings from my analysis reveal that majority of account closures are happening among the customers with an age group between 35 to 55.It is also worth noticing that these are the age groups with more active accounts.When it comes to the employment status,48.5% of customers who close their accounts are working part-time followed by full-time employees with 41% despite having 60% and 20% of active accounts respectively.The median income of these people is around 40,000 to 50,000 USD and is not explaining anything related to the customer status.The same applies to customers with different marital status customers.While the spending ratio of two different customers is around 0.7 ;the transaction ratio of customers who are closing their account is 0.18 lower than that of active customers(0.75).The total transactions of closed account customers are at 43 while this number is 71 for active users.It is surprising to observe that the number of dependents doesn't seem to have any effect on the status of the customer.Finally the type of credit which a customer uses clearly distinguishes the active accounts from the closed ones. 58% of account closures are from blue cards users followed by gold card and silver card users with 32.9% and 25.3% respectively.Finally,the total number of blue cards customers constitute about 70% of overall closed accounts.

# Best Classification Model
In order to predict the likelihood of account closures, models such as Logistic Regression,KNN and Random Forest models has been implemented.Among these three models,Random Forest model performed so well on the unforeseen data.This model was able to predict the a customer will close his/her account 93.8% of the time leaving an error rate of 7.2%.It is also capable of distinguishing active and closed account customers 98.5% of time while the other models were struggling at 94% and 93% respectively.The model also tells that total amount spend last year,total transactions,transaction ratio,utilization ratio and spend ratio are the important variables in determining the target customer status.

# Recommendations

Based on the results from my analysis I would recommend bank executives to consider following steps in order to reduce the number of customers from closing their credit card accounts.Since,customers between ages 35-55 are more prone to close their accounts the bank officials should target these groups by understanding their problems through continuous feedback.Since,there is 90% of account closures are from  full-time and part-time customers ,there is a high possibility that they are dissatisfied with the banking functions such as delay in payroll processing/check deposits etc.To overcome this, the bank account should frequency reach out to them through telephonic surveys with proper questionnaires.Since majority of account closures are from blue cards users when compared to other card holders they should work on them credit card offers in the form of discounts and cash backs.The banks should periodically track the customer transaction and spending ratio and observe abnormalities in the banking activity.Based on the predictions obtained from the model,the banks should focus on the customers who are likely to close their accounts and work on them by re-engineering their sales, service, and marketing strategies.If the banks were able to successfully work on all these recommendations they can improve their customer retention and overall brand reputation.



